Synthesis of Linear Functional Programs

### 2021-07-11 article/article.tex
Write polymorphism

We begin with the right invertible rules:

\begin{mathpar}

@@

When we reach a non-invertible proposition on the right, we start inverting the
$\Omega$ context. The rule to transition to inversion on the left is:

\begin{mathpar}
    \infer*[right=($\Uparrow R$)]
    {\Gamma ; \Delta/ \Delta' ; \Omega \Uparrow\ \vdash C \and C\ \textrm{not
    right asynchronous}}
    {\Gamma ; \Delta/\Delta' ; \Omega \vdash C \Uparrow}
\end{mathpar}

Followed by the left invertible rules:

% * L
\begin{mathpar}

@@

When we find a synchronous (i.e. non-invertible) proposition in $\Omega$,
we simply move it to the linear context $\Delta$, and keep inverting on the left:

\begin{mathpar}
    \infer*[right=($\Uparrow L$)]
    {\Gamma; \Delta, A/\Delta'; \Omega \Uparrow\ \vdash C \and A\ 
    \textrm{not left asynchronous}}
    {\Gamma; \Delta/\Delta'; \Omega, A \Uparrow\ \vdash C}
\end{mathpar}

After inverting all the asynchronous propositions in $\Omega$ we'll reach a state
where there are no more propositions to invert ($\Gamma'; \Delta'; \cdot
\Uparrow\ \vdash C$). At this point, we want to \emph{focus} on a proposition.
The focus object can be any of the three: the proposition on the right (the
goal), a proposition from the linear $\Delta$ context, or a proposition from the
unrestricted $\Gamma$ context. For these options we have three \emph{decision}
rules:

\begin{mathpar}
    \infer*[right=(decideR)]
    {\Gamma; \Delta/\Delta' \vdash C \Downarrow \and C\ \textrm{not atomic}}
    {\Gamma; \Delta/\Delta';\cdot \Uparrow\ \vdash C}
\and
    \infer*[right=(decideL)]
    {\Gamma; \Delta/\Delta' ; A \Downarrow\ \vdash C}
    {\Gamma; \Delta, A/\Delta';\cdot \Uparrow\ \vdash C}
\and
    \infer*[right=(decideL!)]
    {\Gamma, A; \Delta/\Delta' ; A \Downarrow\ \vdash C}
    {\Gamma, A; \Delta/\Delta';\cdot \Uparrow\ \vdash C}
\end{mathpar}

The decision rules are naturally followed by either left or right focus rules:

\begin{mathpar}
    \infer*[right=($\tensor R$)]
    {\Gamma; \Delta/\Delta' \vdash M : A \Downarrow \and \Gamma ; \Delta'/\Delta'' \vdash N
    : B \Downarrow}
    {\Gamma; \Delta/\Delta'' \vdash (M \tensor N) : A \tensor B \Downarrow}
\and
    \infer*[right=($1 R$)]
    {  }
    {\Gamma; \Delta/\Delta \vdash \star : 1 \Downarrow}
\and
    \infer*[right=($\oplus R_1$)]
    {\Gamma; \Delta/\Delta' \vdash M : A \Downarrow}
    {\Gamma; \Delta/\Delta' \vdash\ \textrm{inl}\ M : A \oplus B \Downarrow}
\and
    \infer*[right=($\oplus R_2$)]
    {\Gamma; \Delta/\Delta' \vdash M : B \Downarrow}
    {\Gamma; \Delta/\Delta' \vdash\ \textrm{inr}\ M : A \oplus B \Downarrow}
\and
    \infer*[right=($\bang R$)]
    {\Gamma; \Delta/\Delta'; \cdot \vdash M : A \Uparrow \and \Delta = \Delta'}
    {\Gamma; \Delta/\Delta \vdash \bang M : \bang A \Downarrow}
\and
%TODO: eu tenho uma assertion que garante que y não sai com Delta', mas agora
%parece-me inutil?
    \infer*[right=($\lolli L$)]
    {\Gamma; \Delta/\Delta'; y{:}B \Downarrow\ \vdash M : C \and \Gamma;
    \Delta'/\Delta''; \cdot \vdash N : A \Uparrow}
    {\Gamma; \Delta/\Delta''; x{:}A \lolli B \Downarrow\ \vdash M\{(x\,N)/y\} : C}
\and
    \infer*[right=($\with L_1$)]
    {\Gamma; \Delta/\Delta'; y{:}A \Downarrow\ \vdash M : C}
    {\Gamma; \Delta/\Delta'; x{:}A \with B \Downarrow\ \vdash M\{(\textrm{fst}\ x)/y\} : C}
\and
    \infer*[right=($\with L_2$)]
    {\Gamma; \Delta/\Delta'; y{:}B \Downarrow\ \vdash M : C}
    {\Gamma; \Delta/\Delta'; x{:}A \with B \Downarrow\ \vdash M\{(\textrm{snd}\ x)/y\} : C}
\end{mathpar}

Eventually, the focus proposition will no longer be synchronous, i.e. it's atomic or
asynchronous. If we're left focused on an atomic proposition we either instance
the goal or fail. Otherwise the left focus is asynchronous and we can start
inverting it. If we're right focused on a proposition that isn't right
synchronous, we switch to inversion as well. Three rules model these conditions:

\begin{mathpar}
    \infer*[right=(init)]
    {  }
    {\Gamma; \Delta/\Delta'; x{:}A \Downarrow\ \vdash x : A}
\and
    \infer*[right=($\Downarrow L$)]
    {\Gamma; \Delta/\Delta'; A \Uparrow\ \vdash C \and A\ \textrm{not atomic and not left synchronous}}
    {\Gamma; \Delta/\Delta'; A \Downarrow\ \vdash C}
\and
    \infer*[right=($\Downarrow R$)]
    {\Gamma; \Delta/\Delta'; \cdot \vdash A \Uparrow}
    {\Gamma; \Delta/\Delta' \vdash A \Downarrow}
\end{mathpar}

The rules written above together make the core of our synthetiser. Ahead, we'll
present new rules that align and build on top of these to synth recursive
programs from more expressive (richer) types.

\mypara{Recursion} The main idea behind synthesis of recursive programs is the
labeling of the main goal and the addition of its type, under that name, to the
unrestricted context. That is, to synth a function of type $A \lolli B$ named
\emph{foo}, the initial judgment can be written as

\begin{mathpar}
    \infer
    {\dots}
    {\Gamma, \textrm{\emph{foo}}{:}A \lolli B; \Delta/\Delta'; \Omega \vdash M :
    A \lolli B \Uparrow}
\end{mathpar}
and, by definition, all subsequent inference rules will have
($\textrm{\emph{foo}}{:}A \lolli B$) in the $\Gamma$ context too.

We can also force the usage of the recursive call by adding it not only to the
unrestricted context, but to the linear one as well.

@@

The semantics of the simple ADT relate to those of the plus ($\oplus$) type --
both are additive disjunctions.  A proof of an ADT requires only a proof of one
of the types that a constructor takes, similar to the way $\oplus$ requires only
proof of either the left or right type it consists of. Analogously, a proof of
$C$ by deconstructing an ADT stipulates that all types that can construct the
ADT can prove $C$, akin to the left rule for the $\oplus$ connective. In effect,
the inference rules for a simple ADT mimic the $\oplus$ rules, for an
arbitrary ($> 1$) number of types, rather than exactly two. Therefore, there's one
left rule for ADTs, and an arbitrary number of right rules, one for each
constructor:

\begin{mathpar}
    \infer*[right=(adtR)]
    {\Gamma; \Delta/\Delta' \vdash M : X_n \Downarrow}
    {\Gamma; \Delta/\Delta' \vdash\ \textrm{Bar}_n \ M : \te{Foo} \Downarrow}
\and
    \mprset{flushleft}
    \infer*[right=(adtL)]
    {
        \Gamma ; \Delta/ \Delta'_1 ; \Omega, y_1{:}X_1 \Uparrow\ \vdash M_1 : C
        \and
        y_1 \notin \Delta'_1
        \\
        \Gamma ; \Delta/ \Delta'_2 ; \Omega, y_2{:}X_2 \Uparrow\ \vdash M_2 : C
        \\
        y_2 \notin \Delta'_2
        \\\\
        \\ \dots
        \\
        \Gamma ; \Delta/ \Delta'_n ; \Omega, y_n{:}X_n \Uparrow\ \vdash M_n : C
        \\
        y_n \notin \Delta'_n
        \\
        \Delta'_1 = \Delta'_2 = \dots = \Delta'_n
    }
    {\Gamma ; \Delta/\Delta'_1 ; \Omega, x{:}\te{Foo} \Uparrow\
    \vdash\ \textrm{case}\ x\ \textrm{of}\ \dots\ |\ \textrm{Bar}_n\ y_n
    \rightarrow M_n : C}
\end{mathpar}

where the ADT \emph{Foo} and its constructors stand for any ADT with form
\[
    \textrm{data Foo} = \textrm{Bar}_1\ X_1\ \vert\ \textrm{Bar}_2\ X_2\ \vert\
\dots\ \vert\ \textrm{Bar}_n\ X_n
\]

A more general (and interesting) formulation of ADTs says an ADT can be
recursive (or "indutively defined"), meaning the type a constructor needs might
be the type the constructor constructs. Seemingly small, this change has a big
impact in the synthesis process. Take, for instance, the ADT defined as
$\textrm{data Foo} = \textrm{Bar Foo}$, the synth goal $\te{Foo}
\lolli C$, and its derivation:

\begin{mathpar}
    \infer*[right=($\lolli R$)]
    {
        \infer*[right=($\Uparrow R$)]
        {
            \infer*[right=(adtL)]
            {
                \infer*[right=(adtL)]
                {
                    \infer*[right=(adtL)]
                    {\dots}
                    {\Gamma; \Delta/\Delta'; \Omega, z{:}\te{Foo} \Uparrow\ \vdash \textrm{case}\ z\ 
                    \textrm{of Bar}\ k \rightarrow \dots : C}
                }
                {\Gamma; \Delta/\Delta'; \Omega, y{:}\te{Foo} \Uparrow\ \vdash \textrm{case}\ y\ 
                \textrm{of Bar}\ z \rightarrow \dots : C}
            }
            {\Gamma; \Delta/\Delta'; \Omega, x{:}\te{Foo} \Uparrow\ \vdash \textrm{case}\ x\ 
            \textrm{of Bar}\ y \rightarrow \dots : C}
        }
        {\Gamma; \Delta/\Delta'; \Omega, x{:}\te{Foo} \vdash \dots : C
        \Uparrow}
    }
    {\Gamma; \Delta/\Delta'; \Omega \vdash \lambda x \rightarrow \dots :
    \te{Foo} \lolli C \Uparrow}
\end{mathpar}

Using our current system, we are to apply an infinite number of times the left
ADT rule, never reaching a final state for the proof. Simetrically, the
derivation for the synth goal $\te{Foo}$ doesn't have an end either.

\begin{mathpar}
    \infer*[left=(adtR)]
    {
        \infer*[left=(adtR)]
        {
            \infer*[left=(adtR)]
            {
                \infer*[left=(adtR)]
                {\dots}
                {\Gamma; \Delta/\Delta'; \Omega \vdash \textrm{Bar}\ \dots : \te{Foo} \Downarrow}
            }
            {\Gamma; \Delta/\Delta'; \Omega \vdash \textrm{Bar}\ \dots : \te{Foo} \Downarrow}
        }
        {\Gamma; \Delta/\Delta'; \Omega \vdash \textrm{Bar}\ \dots : \te{Foo} \Downarrow}
    }
    {\Gamma; \Delta/\Delta'; \Omega \vdash \textrm{Bar}\ \dots : \te{Foo}
    \Downarrow}
\end{mathpar}

To handle this situation, we impede the decomposition of an ADT in subsequent
proofs of its branches, and to impede the construction of an ADT in a subsequent
proof of a type to construct it. For this, we need two more contexts, $\Rho_C$
for restrictions on construction and $\Rho_D$ for restrictions on
deconstruction. For convenience, they are represented by a single $\Rho$ if
unused. All rules must trivially propagate these. The ADT rules are explicitly
modified to take this into consideration:

\begin{mathpar}
    \infer*[right=(adtR)]
    {(\Rho_C'; \Rho_D) ; \Gamma; \Delta/\Delta' \vdash M : X_n \Downarrow \and
    \te{Foo} \notin \Rho_C}
    {(\Rho_C; \Rho_D);\Gamma; \Delta/\Delta' \vdash\ \textrm{Bar}_n \ M : \te{Foo} \Downarrow}
\and
    \mprset{flushleft}
    \infer*[right=(adtL)]
    {
        \te{Foo} \notin \Rho_D
        \and
        \Delta'_1 = \dots = \Delta'_n 
        \\
        (\Rho_C; \Rho'_D);\Gamma ; \Delta/ \Delta'_1 ; \Omega, y_1{:}X_1 \Uparrow\ \vdash M_1 : C
        \and
        y_1 \notin \Delta'_1
        \\\\
        \\ \dots
        \\\\
        (\Rho_C; \Rho'_D);\Gamma ; \Delta/ \Delta'_n ; \Omega, y_n{:}X_n \Uparrow\ \vdash M_n : C
        \and
        y_n \notin \Delta'_n
    }
    {(\Rho_C; \Rho_D); \Gamma ; \Delta/\Delta'_1 ; \Omega, x{:}\te{Foo} \Uparrow\
    \vdash\ \textrm{case}\ x\ \textrm{of}\ \dots\ |\ \textrm{Bar}_n\ y_n
    \rightarrow M_n : C}
\end{mathpar}
where
\begin{mathpar}
    \Rho'_C = \textrm{\textbf{if}}\ \te{Foo}\ \textrm{is recursive \textbf{then}}\ \Rho_C,
    \te{Foo}\ \textrm{\textbf{else}}\ \Rho_C

    \Rho'_D = \textrm{likewise}
\end{mathpar}

These modifications prevent the derivation from not ending for the two seen
situations, however, they also greatly limit the derivable valid programs space,
leaving the synthetiser pretty much unable to synth specifications with
recursive types, which is so undesirable that we add two rules to complement the
restrictions on construction and destruction of recursive types.

First, after decomposing a recursive ADT, we must consume all constructor
parameters, and since we can't decompose the self type any further because of a
restriction, we should be able to use it from the linear context; even though,
non-restricted types are not to take that course since they weren't unnaturally
restricted, and because it would result in additional non-determinism and
infinite recursion. Second, without any additional rules, an ADT in the linear
context will loop back to the inversion context, and then loop again, going back and
forth between the two contexts for infinity; instead, when focused on an ADT, we should
either instance the goal (provided they're the same type), fail if the
decomposition is restricted, or switch to inversion if it's not. The three
following rules ensure this:

\begin{mathpar}
    \infer*[right=(adt$\Uparrow$L)]
    {
        (\Rho_C; \Rho_D);\Gamma; \Delta, x{:}\te{Foo}/\Delta'; \Omega \Uparrow\ \vdash M : C
        \and
        \te{Foo} \in \Rho_D
    }
    {(\Rho_C; \Rho_D);\Gamma; \Delta/\Delta'; \Omega, x{:}\te{Foo} \Uparrow\ \vdash M : C}
    \and
    \infer*[right=(adt-init)]
    {  }
    {\Rho; \Gamma; \Delta/\Delta'; x{:}\te{Foo} \Downarrow\ \vdash x : \te{Foo}}
    \and
    \infer*[right=(adt$\Downarrow$L)]
    {(\Rho_C; \Rho_D); \Gamma; \Delta/\Delta'; x{:}\te{Foo} \Uparrow\ \vdash M :
    \te{Foo} \and \te{Foo} \notin \Rho_D}
    {(\Rho_C; \Rho_D); \Gamma; \Delta/\Delta'; x{:}\te{Foo} \Downarrow\ \vdash M : \te{Foo}}
\end{mathpar}

% TODO: Adicionar regra que diz que se for ADT Rec -o A então a construção de
% ADT Rec é logo restrita?? parece que acelera minimamente mas não arranjo um
% exemplo em que seja crucial. pode ser uma coisa que tenha ficado para resolver
% uma coisa antiga que ficou resolvida mais tarde com outra modificação e então
% é agora inutil. Talvez seja demais adicionar à regra

conclude...

\mypara{Polimorfic Types}

>>>>

We begin with the right invertible rules:
\begin{mathpar}

@@

When we reach a non-invertible proposition on the right, we start inverting the
$\Omega$ context. The rule to transition to inversion on the left is:
\begin{mathpar}
    \infer*[right=($\Uparrow R$)]
    {\Gamma ; \Delta/ \Delta' ; \Omega \Uparrow\ \vdash C \and C\ \textrm{not
    right asynchronous}}
    {\Gamma ; \Delta/\Delta' ; \Omega \vdash C \Uparrow}
\end{mathpar}

Followed by the left invertible rules:
\begin{mathpar}

@@

When we find a synchronous (i.e. non-invertible) proposition in $\Omega$,
we simply move it to the linear context $\Delta$, and keep inverting on the left:
\begin{mathpar}
    \infer*[right=($\Uparrow L$)]
    {\Gamma; \Delta, A/\Delta'; \Omega \Uparrow\ \vdash C \and A\ 
    \textrm{not left asynchronous}}
    {\Gamma; \Delta/\Delta'; \Omega, A \Uparrow\ \vdash C}
\end{mathpar}

After inverting all the asynchronous propositions in $\Omega$ we'll reach a state
where there are no more propositions to invert ($\Gamma'; \Delta'; \cdot
\Uparrow\ \vdash C$). At this point, we want to \emph{focus} on a proposition.
The focus object can be any of the three: the proposition on the right (the
goal), a proposition from the linear $\Delta$ context, or a proposition from the
unrestricted $\Gamma$ context. For these options we have three \emph{decision}
rules:
\begin{mathpar}
    \infer*[right=(decideR)]
    {\Gamma; \Delta/\Delta' \vdash C \Downarrow \and C\ \textrm{not atomic}}
    {\Gamma; \Delta/\Delta';\cdot \Uparrow\ \vdash C}
\and
    \infer*[right=(decideL)]
    {\Gamma; \Delta/\Delta' ; A \Downarrow\ \vdash C}
    {\Gamma; \Delta, A/\Delta';\cdot \Uparrow\ \vdash C}
\and
    \infer*[right=(decideL!)]
    {\Gamma, A; \Delta/\Delta' ; A \Downarrow\ \vdash C}
    {\Gamma, A; \Delta/\Delta';\cdot \Uparrow\ \vdash C}
\end{mathpar}

The decision rules are naturally followed by either left or right focus rules:
\begin{mathpar}
    \infer*[right=($\tensor R$)]
    {\Gamma; \Delta/\Delta' \vdash M : A \Downarrow \and \Gamma ; \Delta'/\Delta'' \vdash N
    : B \Downarrow}
    {\Gamma; \Delta/\Delta'' \vdash (M \tensor N) : A \tensor B \Downarrow}
\and
    \infer*[right=($1 R$)]
    {  }
    {\Gamma; \Delta/\Delta \vdash \star : 1 \Downarrow}
\and
    \infer*[right=($\oplus R_1$)]
    {\Gamma; \Delta/\Delta' \vdash M : A \Downarrow}
    {\Gamma; \Delta/\Delta' \vdash\ \textrm{inl}\ M : A \oplus B \Downarrow}
\and
    \infer*[right=($\oplus R_2$)]
    {\Gamma; \Delta/\Delta' \vdash M : B \Downarrow}
    {\Gamma; \Delta/\Delta' \vdash\ \textrm{inr}\ M : A \oplus B \Downarrow}
\and
    \infer*[right=($\bang R$)]
    {\Gamma; \Delta/\Delta'; \cdot \vdash M : A \Uparrow \and \Delta = \Delta'}
    {\Gamma; \Delta/\Delta \vdash \bang M : \bang A \Downarrow}
\and
%TODO: eu tenho uma assertion que garante que y não sai com Delta', mas agora
%parece-me inutil?
    \infer*[right=($\lolli L$)]
    {\Gamma; \Delta/\Delta'; y{:}B \Downarrow\ \vdash M : C \and \Gamma;
    \Delta'/\Delta''; \cdot \vdash N : A \Uparrow}
    {\Gamma; \Delta/\Delta''; x{:}A \lolli B \Downarrow\ \vdash M\{(x\,N)/y\} : C}
\and
    \infer*[right=($\with L_1$)]
    {\Gamma; \Delta/\Delta'; y{:}A \Downarrow\ \vdash M : C}
    {\Gamma; \Delta/\Delta'; x{:}A \with B \Downarrow\ \vdash M\{(\textrm{fst}\ x)/y\} : C}
\and
    \infer*[right=($\with L_2$)]
    {\Gamma; \Delta/\Delta'; y{:}B \Downarrow\ \vdash M : C}
    {\Gamma; \Delta/\Delta'; x{:}A \with B \Downarrow\ \vdash M\{(\textrm{snd}\ x)/y\} : C}
\end{mathpar}
Eventually, the focus proposition will no longer be synchronous, i.e. it's atomic or
asynchronous. If we're left focused on an atomic proposition we either instance
the goal or fail. Otherwise the left focus is asynchronous and we can start
inverting it. If we're right focused on a proposition that isn't right
synchronous, we switch to inversion as well. Three rules model these conditions:
\begin{mathpar}
    \infer*[right=(init)]
    {  }
    {\Gamma; \Delta/\Delta'; x{:}A \Downarrow\ \vdash x : A}
\and
    \infer*[right=($\Downarrow L$)]
    {\Gamma; \Delta/\Delta'; A \Uparrow\ \vdash C \and A\ \textrm{not atomic and not left synchronous}}
    {\Gamma; \Delta/\Delta'; A \Downarrow\ \vdash C}
\and
    \infer*[right=($\Downarrow R$)]
    {\Gamma; \Delta/\Delta'; \cdot \vdash A \Uparrow}
    {\Gamma; \Delta/\Delta' \vdash A \Downarrow}
\end{mathpar}
The rules written above together make the core of our synthetiser. Ahead, we'll
present new rules that align and build on top of these to synth recursive
programs from more expressive (richer) types.

\mypara{Recursion} The main idea behind synthesis of recursive programs is the
labeling of the main goal and the addition of its type, under that name, to the
unrestricted context. That is, to synth a function of type $A \lolli B$ named
\emph{foo}, the initial judgment can be written as
\begin{mathpar}
    \infer
    {\dots}
    {\Gamma, \textrm{\emph{foo}}{:}A \lolli B; \Delta/\Delta'; \Omega \vdash M :
    A \lolli B \Uparrow}
\end{mathpar}
and, by definition, all subsequent inference rules will have
($\textrm{\emph{foo}}{:}A \lolli B$) in the $\Gamma$ context too.
We can also force the usage of the recursive call by adding it not only to the
unrestricted context, but to the linear one as well.

@@

The semantics of the simple ADT relate to those of the plus ($\oplus$) type --
both are additive disjunctions.  A proof of an ADT requires only a proof of one
of the types that a constructor takes, similar to the way $\oplus$ requires only
proof of either the left or right type it consists of. Analogously, a proof of
$C$ by deconstructing an ADT stipulates that all types that can construct the
ADT can prove $C$, akin to the left rule for the $\oplus$ connective. In effect,
the inference rules for a simple ADT mimic the $\oplus$ rules, for an
arbitrary ($> 1$) number of types, rather than exactly two. Therefore, there's one
left rule for ADTs, and an arbitrary number of right rules, one for each
constructor:
\begin{mathpar}
    \infer*[right=(adtR)]
    {\Gamma; \Delta/\Delta' \vdash M : X_n \Downarrow}
    {\Gamma; \Delta/\Delta' \vdash\ \textrm{Bar}_n \ M : \te{Foo} \Downarrow}
\and
    \mprset{flushleft}
    \infer*[right=(adtL)]
    {
        \Gamma ; \Delta/ \Delta'_1 ; \Omega, y_1{:}X_1 \Uparrow\ \vdash M_1 : C
        \and
        y_1 \notin \Delta'_1
        \\
        \Gamma ; \Delta/ \Delta'_2 ; \Omega, y_2{:}X_2 \Uparrow\ \vdash M_2 : C
        \\
        y_2 \notin \Delta'_2
        \\\\
        \\ \dots
        \\
        \Gamma ; \Delta/ \Delta'_n ; \Omega, y_n{:}X_n \Uparrow\ \vdash M_n : C
        \\
        y_n \notin \Delta'_n
        \\
        \Delta'_1 = \Delta'_2 = \dots = \Delta'_n
    }
    {\Gamma ; \Delta/\Delta'_1 ; \Omega, x{:}\te{Foo} \Uparrow\
    \vdash\ \textrm{case}\ x\ \textrm{of}\ \dots\ |\ \textrm{Bar}_n\ y_n
    \rightarrow M_n : C}
\end{mathpar}
where the ADT \emph{Foo} and its constructors stand for any ADT with form
\[
    \textrm{data Foo} = \textrm{Bar}_1\ X_1\ \vert\ \textrm{Bar}_2\ X_2\ \vert\
\dots\ \vert\ \textrm{Bar}_n\ X_n
\]

A more general (and interesting) formulation of ADTs says an ADT can be
recursive (or "indutively defined"), meaning the type a constructor needs might
be the type the constructor constructs. Seemingly small, this change has a big
impact in the synthesis process. Take, for instance, the ADT defined as
$\textrm{data Foo} = \textrm{Bar Foo}$, the synth goal $\te{Foo}
\lolli C$, and its derivation:
\begin{mathpar}
    \infer*[right=($\lolli R$)]
    {
        \infer*[right=($\Uparrow R$)]
        {
            \infer*[right=(adtL)]
            {
                \infer*[right=(adtL)]
                {
                    \infer*[right=(adtL)]
                    {\dots}
                    {\Gamma; \Delta/\Delta'; \Omega, z{:}\te{Foo} \Uparrow\ \vdash \textrm{case}\ z\ 
                    \textrm{of Bar}\ k \rightarrow \dots : C}
                }
                {\Gamma; \Delta/\Delta'; \Omega, y{:}\te{Foo} \Uparrow\ \vdash \textrm{case}\ y\ 
                \textrm{of Bar}\ z \rightarrow \dots : C}
            }
            {\Gamma; \Delta/\Delta'; \Omega, x{:}\te{Foo} \Uparrow\ \vdash \textrm{case}\ x\ 
            \textrm{of Bar}\ y \rightarrow \dots : C}
        }
        {\Gamma; \Delta/\Delta'; \Omega, x{:}\te{Foo} \vdash \dots : C
        \Uparrow}
    }
    {\Gamma; \Delta/\Delta'; \Omega \vdash \lambda x \rightarrow \dots :
    \te{Foo} \lolli C \Uparrow}
\end{mathpar}
Using our current system, we are to apply an infinite number of times the left
ADT rule, never reaching a final state for the proof. Simetrically, the
derivation for the synth goal $\te{Foo}$ doesn't have an end either.
\begin{mathpar}
    \infer*[left=(adtR)]
    {
        \infer*[left=(adtR)]
        {
            \infer*[left=(adtR)]
            {
                \infer*[left=(adtR)]
                {\dots}
                {\Gamma; \Delta/\Delta'; \Omega \vdash \textrm{Bar}\ \dots : \te{Foo} \Downarrow}
            }
            {\Gamma; \Delta/\Delta'; \Omega \vdash \textrm{Bar}\ \dots : \te{Foo} \Downarrow}
        }
        {\Gamma; \Delta/\Delta'; \Omega \vdash \textrm{Bar}\ \dots : \te{Foo} \Downarrow}
    }
    {\Gamma; \Delta/\Delta'; \Omega \vdash \textrm{Bar}\ \dots : \te{Foo}
    \Downarrow}
\end{mathpar}
To handle this situation, we impede the decomposition of an ADT in subsequent
proofs of its branches, and to impede the construction of an ADT in a subsequent
proof of a type to construct it. For this, we need two more contexts, $\Rho_C$
for restrictions on construction and $\Rho_D$ for restrictions on
deconstruction. For convenience, they are represented by a single $\Rho$ if
unused. All rules must trivially propagate these. The ADT rules are explicitly
modified to take this into consideration:
\begin{mathpar}
    \infer*[right=(adtR)]
    {(\Rho_C'; \Rho_D) ; \Gamma; \Delta/\Delta' \vdash M : X_n \Downarrow \and
    \te{Foo} \notin \Rho_C}
    {(\Rho_C; \Rho_D);\Gamma; \Delta/\Delta' \vdash\ \textrm{Bar}_n \ M : \te{Foo} \Downarrow}
\and
    \mprset{flushleft}
    \infer*[right=(adtL)]
    {
        \te{Foo} \notin \Rho_D
        \and
        \Delta'_1 = \dots = \Delta'_n 
        \\
        (\Rho_C; \Rho'_D);\Gamma ; \Delta/ \Delta'_1 ; \Omega, y_1{:}X_1 \Uparrow\ \vdash M_1 : C
        \and
        y_1 \notin \Delta'_1
        \\\\
        \\ \dots
        \\\\
        (\Rho_C; \Rho'_D);\Gamma ; \Delta/ \Delta'_n ; \Omega, y_n{:}X_n \Uparrow\ \vdash M_n : C
        \and
        y_n \notin \Delta'_n
    }
    {(\Rho_C; \Rho_D); \Gamma ; \Delta/\Delta'_1 ; \Omega, x{:}\te{Foo} \Uparrow\
    \vdash\ \textrm{case}\ x\ \textrm{of}\ \dots\ |\ \textrm{Bar}_n\ y_n
    \rightarrow M_n : C}
\end{mathpar}
where
\begin{mathpar}
    \Rho'_C = \textrm{\textbf{if}}\ \te{Foo}\ \textrm{is recursive \textbf{then}}\ \Rho_C,
    \te{Foo}\ \textrm{\textbf{else}}\ \Rho_C

    \Rho'_D = \textrm{likewise}
\end{mathpar}
These modifications prevent the derivation from not ending for the two seen
situations, however, they also greatly limit the derivable valid programs space,
leaving the synthetiser pretty much unable to synth specifications with
recursive types, which is so undesirable that we add three rules to complement the
restrictions on construction and destruction of recursive types.

First, after decomposing a recursive ADT, we must consume all constructor
parameters, and since we can't decompose the self type any further because of a
restriction, we should be able to use it from the linear context; even though,
non-restricted types are not to take that course since they weren't unnaturally
restricted, and because it would result in additional non-determinism and
infinite recursion. Second, without any additional rules, an ADT in the linear
context will loop back to the inversion context, and then loop again, going back and
forth between the two contexts for infinity; instead, when focused on an ADT, we should
either instance the goal (provided they're the same type), or switch to
inversion if and only if its decomposition isn't restricted. The three
following rules ensure this:
\begin{mathpar}
    \infer*[right=(adt$\Uparrow$L)]
    {
        (\Rho_C; \Rho_D);\Gamma; \Delta, x{:}\te{Foo}/\Delta'; \Omega \Uparrow\ \vdash M : C
        \and
        \te{Foo} \in \Rho_D
    }
    {(\Rho_C; \Rho_D);\Gamma; \Delta/\Delta'; \Omega, x{:}\te{Foo} \Uparrow\ \vdash M : C}
    \and
    \infer*[right=(adt-init)]
    {  }
    {\Rho; \Gamma; \Delta/\Delta'; x{:}\te{Foo} \Downarrow\ \vdash x : \te{Foo}}
    \and
    \infer*[right=(adt$\Downarrow$L)]
    {(\Rho_C; \Rho_D); \Gamma; \Delta/\Delta'; x{:}\te{Foo} \Uparrow\ \vdash M :
    \te{Foo} \and \te{Foo} \notin \Rho_D}
    {(\Rho_C; \Rho_D); \Gamma; \Delta/\Delta'; x{:}\te{Foo} \Downarrow\ \vdash M : \te{Foo}}
\end{mathpar}
% TODO: Adicionar regra que diz que se for ADT Rec -o A então a construção de
% ADT Rec é logo restrita?? parece que acelera minimamente mas não arranjo um
% exemplo em que seja crucial. pode ser uma coisa que tenha ficado para resolver
% uma coisa antiga que ficou resolvida mais tarde com outra modificação e então
% é agora inutil. Talvez seja demais adicionar à regra
The ADT rules draw closer to a final form, but will still need a revision
after the introduction of polymorfism and polymorfic parametrization on ADTs.

\mypara{Polimorfic Types} A polymorfic specification is a type $X$ and a set of
bound variables that quantify universally the type variables in $X$, called a
\emph{scheme} ($\forall \overline{\alpha}.\ \tau$). Synthesis of a scheme is straightforward.
First, \textbf{type variables are considered atomic types}, then, instantiate
the bound vars of the scheme as described by the Hindley-Milner inference
method's~\cite{HM-infer} instantiation rule (put simply, generate fresh names for
each bound type variable); e.g. the scheme $\forall \alpha.\ \alpha \lolli \alpha$ could be instanced to
$\alpha0 \lolli \alpha0$. We add a rule for this:
\begin{mathpar}
    \infer*[right=($\forall R$)]
    { \Rho; \Gamma; \Delta/\Delta'; \Omega \vdash \tau' \Uparrow \and \forall
    \overline{\alpha}.\ \tau
    \sqsubseteq \tau'}
    {\Rho; \Gamma; \Delta/\Delta'; \Omega \vdash \forall \overline{\alpha}.\
    \tau \Uparrow}
\end{mathpar}
where
\[
    \forall \overline{\alpha}.\ \tau \sqsubseteq \tau'\ \textrm{indicates type}\
    \tau'\ \textrm{is an \emph{instantiation} of type scheme}\ \forall
    \overline{\alpha}.\ \tau
\]
As such, the construction of a derivation in which, to some types, no specific rules can be applied (except for the \textsc{init} rule,
a type variable can instance a goal that's a type variable of the same name),
corresponds to the synthesis of a program where some expressions are treated
agnostically -- nothing constrains their type, meaning the program is
polymorphic * TODO: rewrite this?.
